Ages of Planets and Stars Observed by Kepler

### 2024-04-02 paper/ms.tex
update ms

%%%%%%%%%
% STARS %
%%%%%%%%%
% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of kepler stars w/ rotation periods
\newcommand{\nstarswithprot}{{$\sim$50{,}000}}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}

@@

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
We summarize the origin of stellar parameters other than the ages in
Section~\ref{sec:stellarprops}, 
and describe and validate our age-dating methods in
Section~\ref{sec:agemethod}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

@@

\subsection{Kepler stars with rotation periods}
\label{subsec:starsel}

We select stars with rotation periods as determined by previous
investigators.  Studies of stellar rotation across the entire Kepler field
include those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015},
\citet{Santos_2019} (\citetalias{Santos_2019}), and
\citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach
based on the autocorrelation function to detect 34{,}030 rotation
periods for main-sequence Kepler targets cooler than 6500\,K, and excluded
known eclipsing binaries and KOIs.  
\citet{Reinhold_2015} used an iterative Lomb-Scargle based approach to
analyze $\approx$40{,}000 stars with a variability range $R_{\rm
	var}>0.3$\% that were not known EBs, planet candidates, or
  pulsators;
they reported primary rotation periods for 24{,}124 of these stars.
\citet{Reinhold_2015} also reported secondary periods for two thirds
of the stars with primary periods, which could be caused by
differential rotation or finite spot lifetimes.  
Finally,
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  
The rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
based on updated knowledge of contaminants (e.g.~giant stars and
eclipsing binaries) and visual inspection.  In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Our speculation is that
these reported detections are more challenging than those from
\citetalias{McQuillan_2014}, perhaps due to a lower variability
amplitude, or a longer rotation period.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.
We verified that the main conclusions of our analysis do not change if
we were to adopt only the periods from \citetalias{McQuillan_2014}.

Concatenating the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} yields \nstarswithprot\ Kepler stars with
rotation periods.

%FIXME

@@

As our default source for stellar temperatures and surface gravities,
we adopt the values from the Gaia-Kepler Stellar Properties Catalog
(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
available for stars with ``AAA'' 2MASS photometry, measured parallaxes
in Gaia DR2,  and $g$-band photometry from either the KIC or the
Kepler-INT survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars
with rotation periods.

@@

\section{Age Measurement}
\label{sec:agemethod}

\subsection{Gyrochrone ages}
To calculate gyrochrone ages, we use \texttt{gyro-interp}
\citep{Bouma_2023}.  This framework was built to statistically address
the issue that, especially over the first gigayear of stellar
spin-down, stars with the same mass and same rotation period can have
very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
gyrochrone age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
measured.  The three main assumptions behind the \texttt{gyro-interp}
framework are as follows.  {\it 1)} The stars are on the main
sequence, {\it 2)} they do not have any binary companions,
and {\it 3)} the rotation period distribution
for field stars of a given mass and age is identical to that of
the calibration open clusters (metallicity differences, for instance, are
ignored).  
Regarding the former two assumptions:
pre-main-sequence and post-main-sequence evolution introduce
variations in the stellar angular momentum budget.  Similarly, tidal
effects might systematically spin up (or down) stars in binaries.

Using our adopted effective temperatures and rotation periods, we ran 
\texttt{gyro-interp} over an age grid of 0 to 5\,Gyr, linearly spaced
over 500 grid points.
\texttt{gyro-interp} itself is only calibrated out to 4\,Gyr, the
adopted age of the M67 cluster \citep[see][]{Dungee_2022,Gruner_2023}.
We used the spin-down law between the reference open clusters by
interpolating using piecewise cubic hermite interpolating polynomials,
and integrated Equation~1 of \citep{Bouma_2023} at the default grid
resolution.
This yielded age posterior probabilities over our requested age grid
for \nkepstarswithages\ Kepler stars, and \nkoisnofpwithages\
non-false-positive KOIs.

\subsection{Lithium ages}
To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ feature is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

\subsection{Quality flags}
In our core gyrochrone analysis, we calculate gyrochrone age
posteriors for all the stars with reported rotation periods and
effective temperatures.  To select the stars for which we expect the
gyrochronal age to be valid, we then build a set of quality flags, and
apply them in the sub-analyses that follow.  We defined these filters
as follows, each of which is included in Tables~\ref{tab:stars}
and~\ref{tab:planets}.

@@

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.8\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
		}
		\vspace{-0.62cm}
		\subfloat{
			\includegraphics[width=0.66\textwidth]{koi_mean_prot_teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
    These periods are from \citet{Santos_2019}
    (\citetalias{Santos_2019}) and \citet{Santos_2021}
    (\citetalias{Santos_2021}).
    The gray models... %TODO FIXME
	}
	\label{fig:prot_vs_teff}
\end{figure*}

>>>>

%%%%%%%%%
% STARS %
%%%%%%%%%
\input{vals_manuscript}

% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}

@@

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
We summarize the origin of stellar parameters other than the ages in
Section~\ref{sec:stellarprops}, 
and describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

@@

\subsection{Rotation periods}
\label{subsec:starsel}

We select stars with rotation periods that have been measured by
previous investigators.  Studies of stellar rotation across the entire
Kepler field include those by \citet{McQuillan_2014}
(\citetalias{McQuillan_2014}), \citet{Reinhold_2015},
\citet{Santos_2019} (\citetalias{Santos_2019}), and
\citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach based on the
autocorrelation function to detect 34{,}030 rotation periods for
main-sequence Kepler targets cooler than 6500\,K, and excluded known
eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
candidates, or pulsators; they reported primary rotation periods for
24{,}124 of these stars.  \citet{Reinhold_2015} also reported
secondary periods for two thirds of the stars with primary periods,
which could be caused by differential rotation or finite spot
lifetimes.  Finally, \citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  The
rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
based on updated knowledge of contaminants (e.g.~giant stars and
eclipsing binaries) and visual inspection.  In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Our speculation is that
these reported detections are more challenging than those from
\citetalias{McQuillan_2014}, perhaps due to a lower variability
amplitude, or a longer rotation period.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.
%TODO is it?
We verified that the main conclusions of our analysis do not change if
we were to adopt only the periods from \citetalias{McQuillan_2014}.
%TODO have you?

Concatenating the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} yields \nuniqstarsantosrot\ Kepler stars with
rotation periods.

@@

As our default source for stellar temperatures and surface gravities,
we adopt the values from the Gaia-Kepler Stellar Properties Catalog
(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
available for stars with ``AAA'' 2MASS photometry, measured parallaxes
in Gaia DR2,  and $g$-band photometry from either the KIC or the
Kepler-INT survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
stars with rotation periods.

@@

\section{Ages From Rotation}
\label{sec:rotage}

We calculate gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}.  We built this statistical framework to address
the issue that, especially over the first gigayear of stellar
spin-down, stars with the same mass and same rotation period can have
very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
gyrochrone age posterior should therefore incorporate this intrinsic
population-level scatter into the precision with which the age can be
measured.  The three main assumptions behind the \texttt{gyro-interp}
framework are as follows.  {\it 1)} The stars are on the main
sequence, {\it 2)} their spindown is not influenced by binary
companions, and {\it 3)} the rotation period distribution for field
stars of a given mass and age is identical to that of the calibration
open clusters (metallicity differences, for instance, are ignored).  
Regarding the former two assumptions: pre-main-sequence and
post-main-sequence evolution introduce variations in the stellar
angular momentum budget.  Similarly, tidal effects might
systematically spin up (or down) stars in binaries.

Using our adopted effective temperatures and rotation periods, we ran 
\texttt{gyro-interp} over an age grid of 0 to 5\,Gyr, linearly spaced
over 500 grid points.
\texttt{gyro-interp} itself is only calibrated out to 4\,Gyr, the
adopted age of the M67 cluster \citep[see][]{Dungee_2022,Gruner_2023}.
We used the spin-down law between the reference open clusters by
interpolating using piecewise cubic hermite interpolating polynomials,
and integrated Equation~1 of \citep{Bouma_2023} at the default grid
resolution.

\subsection{Star Filters}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of filters and quality flags, and apply
them in the sub-analyses that follow.  We defined these filters as
follows, each of which is included in Tables~\ref{tab:stars}
and~\ref{tab:planets}.

@@

\subsection{Planet Filters}
For the subset of our analysis that focuses on the properties of
planets, we use the following additional filters.

% see Erik's 2022 discussion; might want to shift to include a cut on
% disposition score.
{\it Candidate reliability}---no false positives; only ``planet
candidates'' and ``confirmed planets''.

{\it Candidate S/N}---the false positive rate increases greatly toward
Kepler's noise floor for planet detection (CITE).  We required a S/N
in excess of the usual floor, through a cut on the ``multiple event
statistic'' (MES): we required ${\rm MES}$$>$10.

{\it Grazing planets}---grazing objects ($b$>0.8) often provide
non-robust planetary sizes due to the size-impact parameter
degeneracy.  Particularly for large planets, they can also include
astrophysical false positives at higher rates (due to the same
degeneracy).  While we include a flag for such objects, we by default
do not exclude them from our analysis; this is because...{\bf justify}.

\subsection{Tally}

The rotation periods for the stellar and planetary samples are shown
in Figure~\ref{fig:prot_vs_teff}.
Out of the \nkeplerstars\ stars observed by Kepler, requiring reported
rotation periods yielded \nuniqstarsantosrotgyroappl.  Further
requiring these rotating stars to be apparently single, near the main
sequence, with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
eff}$/K$<$6200, and not flagged as eclipsing binaries yielded
\nuniqstarsantosrotgyroappl\ rotators amenable for gyrochronology.

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Applying all the same stellar cuts, and
requiring these confirmed and candidate planets to have MES$>$10,
yields \nplwgyroage\ planets with rotation-based ages orbiting
\nplhoststarwgyroage\ stars.  If we additionally require the planets
to be non-grazing ($b<0.8$) based on the Markov Chain Monte Carlo
(MCMC) fits from \citet{Thompson_2018}, we are left with
\nplwgyroagenograzing\ and \nplhoststarwgyroagenograzing\ planets and
stars, respectively.

\section{Ages From Lithium}
\label{sec:liage}

To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ feature is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
		}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
    rotation sequences of open clusters.
    The stellar sample shows only apparently single stars near the
    main sequence with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
    eff}$/K$<$6200, that are not flagged as eclipsing binaries.  The
    planet sample has the same stellar cuts, and shows confirmed and
    candidate planets from the cumulative KOI table, requiring them to
    have a MES$>$10.
    % koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png
    % minus
    % koi_mean_prot_teff_koi_X_S19S21dquality_dropgrazing.png
    %Grazing objects ($b$>0.8) include 213 of the planet candidates
    %displayed, and 136 of the host stars.
	}
	\label{fig:prot_vs_teff}
\end{figure*}
